---
title: "4: Coding Basics"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## LESSON OBJECTIVES
1. Develop familiarity with the form and function of the RStudio interface.
2. Apply basic functionality of R
3. Evaluate how basic practies in R contribute to best management practices for data analysis

## DATA TYPES IN R
R treats objects differently based on their characteristics. For more information, please see: https://www.statmethods.net/input/datatypes.html. 

* **Vectors** 1 dimensional structure that contains elements of the same type.
##Can't have different things in vectors (same type of element like numbers)

* **Matrices** 2 dimensional structure that contains elements of the same type.
##Again, same element types like numbers or letters

* **Arrays** Similar to matrices, but can have more than 2 dimensions. We will not delve into arrays in depth.

* **Lists** Ordered collection of elements that can have different modes.
##Can have different data types (a list could contain a vector, an array, etc)


* **Data Frames** 2 dimensional structure that is more general than a matrix. Columns can have different modes (e.g., numeric and factor). When we import csv files into the R workspace, they will enter as data frames.

Define what each new piece of syntax does below (i.e., fill in blank comments). Note that the R chunk has been divided into sections (# at beginning of line, ---- at end)
```{r}
# Vectors ----
vector1 <- c(1,2,5.3,6,-2,4) # numeric vector
vector1

vector2 <- c("one","two","three") # character vector
vector2

vector3 <- c(TRUE,TRUE,TRUE,FALSE,TRUE,FALSE) #logical vector
vector3

vector1[3] # Matrix Subsetting-Returns the third argument in Vector 1



# Matrices ----
matrix1 <- matrix(1:20, nrow = 5,ncol = 4) # Create a matrix with 20 elements with 5 rows and 4 columns. Populates the matrix by column
matrix1

matrix2 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE) #The byrow=TRUE argument populates the matrix horizontally by row. R populates matrices by column automatically as byrow=FALSE (don't need to include this argument).
matrix2


matrix3 <- matrix(1:20, nrow = 5, ncol = 4, byrow = TRUE, # return after comma continues the line
                  dimnames = list(c("uno", "dos", "tres", "cuatro", "cinco"), 
                                  c("un", "deux", "trois", "cat"))) #
matrix3
matrix3["uno",]


matrix1[4, ] #subsetting 4th row of matrix1 and all columns

matrix1[ , 3] #subsetting third column of matrix1 and all rows

matrix1[c(12, 14)] #subsetting values 12 and 14 from matrix1-not including a c would ask R to give us row 12 and column 14, which we don't have. C means concatenate.

matrix1[c(12:14)] #subsetting values 12, 13, and 14 from matrix1


matrix1[2:4, 1:3] #subsetting rows 2:4 and columns 1:3 from matrix1
matrix1[, 1:3] #subsetting all of the rows and columns 1:3 from matrix1



cells <- c(1, 26, 24, 68)
rnames <- c("R1", "R2")      ##create vector called rnames
cnames <- c("C1", "C2")      ##create vector called cnames



matrix4 <- matrix(cells, nrow = 2, ncol = 2, byrow = TRUE,
  dimnames = list(rnames, cnames)) # Create matrix with 2 rows and 2 columns, populating it by row, and name the dimensions as a list of row names and column names
matrix4




# Lists ---- 
list1 <- list(name = "Maria", mynumbers = vector1, mymatrix = matrix1, age = 5.3); list1     ##the semi-colon is a shortcut to print list1. All of the elements within the list are printed out.

list1[[2]]   ##indexing the second element from the list, which returns the mynumbers vector. FOR LISTS, WE USE DOUBLE SQUARE BRACKETS INSTEAD OF USING SINGLE SQUARE BRACKETS





# Data Frames ----
d <- c(1, 2, 3, 4) # Numeric
e <- c("red", "white", "red", NA) # Character
f <- c(TRUE, TRUE, TRUE, FALSE) # Logical

dataframe1 <- data.frame(d,e,f) #tells data frame to combine d, e, and f vectors 
names(dataframe1) <- c("ID","Color","Passed"); View(dataframe1) # Renaming vectors, and View(dataframe1) prints data frame

##NA represents a missing value

dataframe1[1:2] # Returns columns 1:2 from data frame. If we had used this same line of code for a matrix, it would have returned the first 2 elements 1 and 2

matrix1[1:2]



dataframe1[c("ID","Passed")] # Returns specific names of rows and columns (returns columns 2 and 3)


dataframe1$Color # indexes and returns the whole Color column

```
QUESTION: How do the different types of data appear in the Environment tab?

> ANSWER: The Data section in the global environment is able to be viewed in a new window
The values section is not able to be viewed in a new window

QUESTION: In the R chunk below, write "dataframe1$". Press `tab` after you type the dollar sign. What happens?

```{r}
dataframe1$
```

> ANSWER: A dropdown menu shows up, giving you the choice to index one of the three columns in the data frame.


QUESTION: What happens when a comment in R is followed by "----"?

> ANSWER:Denotes a new section. 

Advanced: Sequential section headers can be created by using at least four -, =, and # characters.

## PACKAGES

The Packages tab in the notebook stores the packages that you have saved in your system. A checkmark next to each package indicates whether the package has been loaded into your current R session. Given that R is an open source software, users can create packages that have specific functionalities, with complicated code "packaged" into a simple commands.

If you want to use a specific package that is not in your library already, you need to install it. You can do this in two ways: 

1. Click the install button in the packages tab. Type the package name, which should autocomplete below (case matters). Make sure to check "intall dependencies," which will also install packages that your new package uses. 

2. Type `install.packages("packagename")` into your R chunk or console. It will then appear in your packages list. You only need to do this once. 

If a package is already installed, you will need to load it every session. You can do this in two ways: 

1. Click the box next to the package name in the Packages tab. 

2. Type `library(packagename)` into your R chunk or console.

2a. The command `require(packagename)` will also load a package, but it will not give any error or warning messages if there is an issue.

**Tips and troubleshooting**

* You may be asked to restart R when installing or updating packages. Feel free to say no, as this will obviously slow your progress. However, if the functionality of your new package isn't working properly, try restarting R as a first step. 

* If asked "Do you want to install from sources the packages which needs compilation?", type `yes` into the console. 

* You should only install packages once on your machine. If you store `install.packages` in your R chunks/scripts, comment these lines out, as below. 

* Update your packages regularly! 


```{r}
# We will use the packages dplyr and ggplot2 regularly. 
#install.packages("dplyr") # comment out install commands, use only when needed and re-comment
#install.packages("ggplot2")

library(dplyr)
library(ggplot2)

# Some packages are umbrellas under which other packages are loaded
#install.packages(tidyverse)
library(tidyverse)
```
What happens in the console when you load a package?

> ANSWER: When objects are masked, that means these functions that are part of previous packages are masked

## FUNCTIONS

You've had some practice with functions with the simple commands you've entered in this lesson and the one previous. The basic form of a function is `functionname()`, and the packages we will use in this class will use these basic forms. However, there may be situations when you will want to create your own function. Below is a description of how to write functions through the metaphor of creating a recipe. Credit for this goes to Isabella R. Ghement (@IsabellaGhement on Twitter). 

Writing a function is like writing a recipe. Your function will need a recipe name (functionname). Your recipe ingredients will go inside the parentheses. The recipe steps and end product go inside the curly brackets.
```{r}
functionname <- function(){
  
}
```



A single ingredient recipe: 
```{r}
# Write the recipe. x is the ingredient in the recipe

recipe1 <- function(x){
  mix <- x*2
  return(mix)
}

# Bake the recipe
simplemeal <- recipe1(5)

# Serve the recipe
simplemeal
```



Two single ingredient recipes, baked at the same time: 
```{r}
recipe2 <- function(x){
  mix1 <- x*2
  mix2 <- x/2
  return(list(mix1 = mix1, #comma indicates we continue onto the next line
              mix2 = mix2))
}

doublesimplemeal <- recipe2(6)
doublesimplemeal
```


Two double ingredient recipes, baked at the same time: 
```{r}
recipe3 <- function(x, f){
  mix1 <- x*f
  mix2 <- x/f
  return(list(mix1 = mix1, #comma indicates we continue onto the next line
              mix2 = mix2))
}

doublecomplexmeal <- recipe3(x = 5, f = 2)
doublecomplexmeal
doublecomplexmeal$mix1
```





Make a recipe based on the ingredients you have
```{r}

##If x<3, then multiply x*2. Else X =/<2, then x/2
recipe4 <- function(x) {
  if(x < 3) {
    x*2
  } 
  else {
    x/2
  }
}


##Three conditions: x<3, x>3, or x=3


recipe5 <- function(x) {
  if(x < 3) {
    x*2
  } 
  else if (x > 3) {
    x/2
  }
  else {
    x
  }
}


meal <- recipe4(4); meal

meal2 <- recipe4(2); meal2

meal3 <- recipe5(3); meal3    #gives me X



##Another way of writing If Else Statement
recipe6 <- function(x){
  ifelse(x<3, x*2, x/2)
  
}

meal4 <- recipe6(4); meal4
meal5 <- recipe6(2); meal5
```

