---
title: "9: Data Visualization"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## LESSON OBJECTIVES
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
setwd("~/Desktop/Environmental Data Analytics/Environmental_Data_Analytics/Data/Processed")
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <- read.csv("NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
PeterPaul.chem.nutrients.summaries <- read.csv("NTL-LTER_Lake_Summaries_PeterPaul_Processed.csv")
EPAair <- read.csv("EPAair_O3_PM25_NC1718_Processed.csv")
attach(EPAair)
attach(PeterPaul.chem.nutrients)

```


##Unite Y, m, and d columns into a "date" column

###I fixed Assignment 4 and used the mutate function instead of separate so it kept the original complete date column in the EPAair datatable
```{r}
###EPAair <- EPAair %>%
 ### unite(date, Y, m, d)
###EPAair$date <- as.Date(EPAair$date, format="%m/%d/%y")


```

###unite(data, col, ..., sep = "_", remove = TRUE)


##Create a "month" column in EPAair datatable from the date column
```{r}
###EPAair<- mutate(EPAair, month = month(date))
```


##Create second EPAair data table by separating "date" column into Year, month, and day columns
```{r}
###EPAair2<- separate(EPAair, date, c("Y", "m", "d"))
```


##Create a new Date column in the PeterPaul.chem.nutrients data table in the proper date format from the SampleDate column
```{r}
setwd("~/Desktop/Environmental Data Analytics/Environmental_Data_Analytics/Data/Processed")
PeterPaul.chem.nutrients$Date<-as.Date(PeterPaul.chem.nutrients$sampledate, format="%m/%d/%y")

PeterPaul.chem.nutrients<- mutate(PeterPaul.chem.nutrients, month = month(Date))
PeterPaul.chem.nutrients<-mutate(PeterPaul.chem.nutrients, day=day(Date))


###PeterPaul.chem.nutrients<-separate(PeterPaul.chem.nutrients, Date, c("Y", "m", "d"))
```
###mutate() adds new variables and preserves existing variables.
###separate() turns a single character column into multiple columns but removes original column.



# GGPLOT

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 


ggplot(df, aes()) + geom

## Geoms
Create the actual graph
Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

## Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

time is always on x-axis

# Plotting continuous variables over time: Scatterplot

## Putting aesthetics in ggplot() function-DOESN'T WORK
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()
```

##Putting aesthetics inside geom_point() function
```{r, warning = FALSE}
O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot)
```


# Fix this code

###MUST PUT aesthetics within geom_point() function 
###Can't put color= within aesthetics, so keep it within geom_point() but not in aesthetics!
###Geom_point can take a lot of information besides aesthetics; color is one of them

###Put aesthetics in geom_point
```{r}
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "Blue")
print(O3plot2)
```
###If you change color to "Ozone", the plot becomes a color gradient based on Ozone and adds a legend.




# Add additional variables

##Shape: Changes shape based on what year the data were collected

###Need to say as.factor(year) because the class of year is an integer, and we want it as a factor
###We can put color within aesthetics because we are assigning color by site name
```{r}
PMplot <- 
  ggplot(EPAair, aes(x = Date, y = PM2.5, shape = as.factor(year), color = Site.Name)) +
  geom_point()
print(PMplot)
```



# Separate plot with facets

###Faceting by Site.Name
```{r}
PMplot.faceted <-
  ggplot(EPAair, aes(x = month, y= PM2.5, shape = as.factor(year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3)

print(PMplot.faceted)
```
###Faceting is used when you have one column with two or more "levels"



###Faceting by Year as rows and Site.Name as columns
```{r}
PMplot.faceted2 <-
  ggplot(EPAair, aes(x = month, y = PM2.5)) +
  geom_point() +
  facet_grid(year~Site.Name)
print(PMplot.faceted2)

```

###Faceting by Site.Name as rows and Years as columns
```{r}
PMplot.faceted3 <-
  ggplot(EPAair, aes(x = month, y = PM2.5)) +
  geom_point() +
  facet_grid(Site.Name~year)
print(PMplot.faceted3)

```


# Filter dataset within plot building

###For ex, if we just wanted to plot the "Blackstone" data, we would put Blackstone within quotation marks within subset function, even though in aesthetics function, you don't need quotations
```{r}
O3plot.Blackstone <- 
  ggplot(subset(EPAair, Site.Name == "Blackstone"), aes(x = Date, y = Ozone)) + 
  geom_point() +
  geom_line()
print(O3plot.Blackstone)
```
###Geom_line is connecting the data points along the x-axis. It's only useful if you want to see what the intervals between the data points look like. 



# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

## 1. 


### Plot surface temperatures (depth=0) by day of year. 
### Color your points by year, and facet by lake in two rows.

```{r}
require(ggplot2)
PeterPaulScatterplot<-ggplot(subset(PeterPaul.chem.nutrients, depth=0), aes(x=day, y=temperature_C)) + geom_point(color=year4) +facet_wrap(vars(lakename), nrow=2) 
print(PeterPaulScatterplot)
```


```{r}
PeterPaulScatterplot<-ggplot(subset(PeterPaul.chem.nutrients, depth=0)) + geom_point(aes(x=day, y=temperature_C), color=year4) +facet_wrap(vars(lakename), nrow=2) 
print(PeterPaulScatterplot)
```



#2. 
## Plot temperature by date. Color your points by depth.
## Change the size of your point to 0.5

```{r}
require(ggplot2)
PeterPaulScatterplot2<-ggplot(PeterPaul.chem.nutrients, aes(x=Date, y=temperature_C,color=depth)) + geom_point(size=0.5) +facet_wrap(vars(lakename), nrow=2) 
print(PeterPaulScatterplot2)
```



# Plotting the relationship between two continuous variables: Scatterplot


## Scatterplot
```{r, warning = FALSE}
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

```


# Adjust axes
```{r}
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) +
  ylim(0, 20)
print(lightvsDOfixed)
```
###More light at the surface of the water; from 0-50 irradiance, there are low oxygen concentrations. 




# Depth in the fields of limnology and oceanography is on a reverse scale


###We normally consider depth as an independent variable because it stays the same (temperature might vary)
###In limnology, depth is actually put on the y-axis
```{r}
tempvsdepth <- ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  geom_point() +
  scale_y_reverse()
print(tempvsdepth)

```
###Scale_y_reverse() function reverses the y-axis





```{r}
NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +   ###colored by depth
  geom_point() +
  geom_smooth(method = lm) +   #Creates a line of best fit. method=lm tells it we need a linear model method-must include  this. The gray band around the blue line is a 95% confidence interval-we are 95% confident that this is the linear relationship of our data. We have so many points clustered together, which is why our CI is so narrow
  geom_abline(aes(slope = 16, intercept = 0), lty=2)   #Allows us to draw a linear line with any specifications. lty gives different line types. The points fall mostly above this line, which means this system is more nitrogen rich with less phosphorus molecules. This is common in Midwestern freshwater systems. 
print(NvsP)

```
###colors don't match the gradient
###The gray points mean there was no actual depth associated with that measurement, but there were N and P measurements for those points


# Exercise: Plot relationships between air quality measurements

# 1. 
# Plot AQI values for ozone by PM2.5, colored by site. 
# Add a line of best fit for the linear regression of these variables.
```{r}
AirPlot <- 
  ggplot(EPAair, aes(x = Ozone, y = PM2.5, color = Site.Name)) +
  geom_point() +
   geom_smooth(method = lm)
print(AirPlot)
```
###Triple Oak site has no ozone measurements (y-values) which is why there are no blue points
###Put aesthetics in main ggplot function

  


# Plotting continuous vs. categorical variables

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
 # Barplot + error bars
PeterPaul.nutrient.summaries <- PeterPaul.nutrients.gathered %>%
  group_by(lakename, nutrient) %>%
  summarise(sd = sd(concentration), 
            mean = mean(concentration))

```
###This looks different than other group_by function example in the past because we're now grouping by two factors: lakename and nutrient
###It'll gives us 5 nutrients for both Peter and Paul Lakes


#Bar Plot with Error Bars
```{r}
Nutrientplot <- 
  ggplot(PeterPaul.nutrients.gathered) +
  geom_bar(aes(x = lakename, y = concentration, fill = as.factor(nutrient)), # why did we use fill?
           position = "dodge", stat = "summary", fun.y = "mean")             # what's happening here?
print(Nutrientplot)
```

###fill defines what colors the bars actually are
###Position="Dodge" ensures the bars are separate instead of stacked
###stat="summary and fun.y="mean" takes the summary statistic of a nutrient, and computes the mean of the nutrients to create the bar height

Barcharts require command requires stat function


```{r}
class(nutrient)  #factor
```



#Boxplot with error bars
```{r}
Nutrientplot2 <- 
  ggplot(PeterPaul.nutrient.summaries, aes(x = lakename, y = mean, fill = as.factor(nutrient))) + #
  geom_bar(stat = "identity", position = "dodge") + # what does the stat command do?
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), # how do we specify error bars?
                 position = "dodge")
print(Nutrientplot2)
```
###Stat="identity"--> If you want the heights of the bars to represent values in the data. For a given lake and nutrient, take the identity of that value.
###Aesthetics are now in the ggplot () main function instead of in geom_bar() function
###Geom_bar still has stat and position arguments

#Geom_errorbar function
###Needs aesthetics like any other geom function
###We want the error bars to be one standard deviation below and above the mean. If we didn't have columns for mean and sd in our data summaries, we would have to formulate it ourselves
###Because our geom_bar function has position="dodge"", the geom_errorbar function should also have position="dodge"

###Issues with this graph: These error bars are massive and it's not possible to have a negative concntration of a nutrient. Environmental graphing is moving away from boxplots with error bars. 





# Are there more effective ways to produce summary stats for categories?

# Box and whiskers plot
```{r}
Nutrientplot3 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"?
print(Nutrientplot3)
```
###Color the boxplots by nutrients. 
###fill=nutrient will only color in the middle of boxplot and not the outliers


# Dot plot
```{r}
Nutrientplot4 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge") #
print(Nutrientplot4)
```
###binwidth on y-axis, and choose value of 1 

###This plots every single value measured for a single variable. The wide spread of dot plots at 0 (density) show that there are a lot of different measurements for the concentration. 


# Violin plot
```{r}
Nutrientplot5 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)
```
##plotted same densities of concentrations as dot plot but as a smooth line, instead of dots





# Frequency polygons

## Using a tidy dataset
```{r}
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "black") +
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "darkgray") +
  geom_freqpoly(aes(x = no23), color = "gray") +
  geom_freqpoly(aes(x = po4), color = "blue") 
print(Nutrientplot6)
```
###All nutrients are in separate columns. it draws 5 separate frequency polygons for each nutrient.


# Using a gathered dataset
```{r}
Nutrientplot7 <-   
  ggplot(PeterPaul.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)
```



# Exercise: Plot distributions of AQI values for EPAair

# 1. 
# Create a bar chart plus standard deviation error bars for PM2.5, divided by year. 
# Create separate bars for each site. 


```{r}
EPAAirSummary<- EPAair %>%
  group_by(Site.Name, year) %>%
  summarise(sd = sd(PM2.5, na.rm=TRUE), 
            mean =mean(PM2.5, na.rm=TRUE))
```


```{r}
Nutrientplot <- 
  ggplot(EPAAirSummary, aes(x = as.factor(year), y = mean, fill =Site.Name)) +
  geom_bar(position = "dodge", stat = "summary")  +
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), position="dodge")


print(Nutrientplot)
```






# 2. 
# Create a new plot that better depicts the distribution of PM2.5 concentrations. 
# Divide your graph by year and site.

```
