
---
title: "14: Time Series Analysis"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## LESSON OBJECTIVES
1. Describe the aspects of hierarchical models, fixed effects, and random effects
2. Choose and justify appropriate statistical models when time is an explanatory variable
3. Apply Mann-Kendall and Seasonal Mann-Kendall to datasets with temporal components

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
setwd("~/Desktop/Environmental Data Analytics/Environmental_Data_Analytics/Data/Processed")
library(tidyverse)
install.packages("trend")
library(trend)


PeterPaul.nutrients <- read.csv("NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
USGS.flow.data <- read.csv("USGS_Site02085000_Flow_Raw.csv")

# Rename columns
colnames(USGS.flow.data) <- c("agency_cd", "site_no", "datetime", 
                              "discharge.max", "discharge.max.approval", 
                              "discharge.min", "discharge.min.approval", 
                              "discharge.mean", "discharge.mean.approval", 
                              "gage.height.max", "gage.height.max.approval", 
                              "gage.height.min", "gage.height.min.approval", 
                              "gage.height.mean", "gage.height.mean.approval")


```

# Set date to date format
```{r}
PeterPaul.nutrients$sampledate <- as.Date(PeterPaul.nutrients$sampledate, 
                                               format = "%Y-%m-%d")
USGS.flow.data$datetime <- as.Date(USGS.flow.data$datetime, 
                              format = "%m/%d/%y")


gabytheme <- theme_bw(base_size = 14) + 
  theme(plot.title=element_text(face="bold", size="16", color="tomato2", hjust=0.5),
        axis.title=element_text(face="bold.italic", size=11, color="black"),
axis.text = element_text(face="bold", size=10, color = "black"), 
panel.background=element_rect(fill="lightgray", color="darkblue"), 
panel.border = element_rect(color = "black", size = 2),
legend.position = "top", legend.background = element_rect(fill="white", color="black"),
            legend.key = element_rect(fill="transparent", color="NA"))
theme_set(gabytheme)
```




## NONPARAMETRIC TREND TESTS

In many environmental datasets (especially climate and hydrology), we might not expect a linear trend in the response variable over time. In this case, we will need to employ a nonparametric test to determine whether there is a monotonic trend (i.e., consistent increase or decrease but not necessarily linear) over time. We will illustrate a few examples of nonparametric trend tests today with the `trend` package. 

A vignette for the `trend` package can be found here: https://cran.r-project.org/web/packages/trend/vignettes/trend.pdf. More details here: https://cran.r-project.org/web/packages/trend/trend.pdf.

We will run a Mann-Kendall and a Seasonal Mann-Kendall test today, but there are additional variants of these tests within the package including a correlated Seasonal Mann-Kendall test, a multivariate Mann-Kendall test, a partial Mann-Kendall test, a partial correlation trend test, and a Cox and Stuart trend test. Look into the documentation for these tests to determine which one is appropriate for your purposes.

# Mann-Kendall Test
A Mann-Kendall test will analyze whether there is a monotonic trend in the response variable over time. Let's use the Mann-Kendall test to investigate whether there is a trend in total phosphorus concentrations in Peter Lake over time. 

# Wrangle our dataset
```{r}
###Choose only surface concentrations at depth=0. 
PeterPaul.nutrients.surface <- 
  PeterPaul.nutrients %>%
 ###Remove lakeid, depth_id, and comments. 
  filter(depth == 0) %>%
  filter(!is.na(tp_ug))  ###Remove NA's from tp_ug column 
```


# Initial visualization of data
```{r}
ggplot(PeterPaul.nutrients.surface, aes(x = sampledate, y = tp_ug, color = lakename)) + 
  geom_point() +
  scale_color_manual(values = c("#7fcdbb", "#253494"))
```

# Split dataset by lake
```{r}
Peter.nutrients.surface <- filter(PeterPaul.nutrients.surface, lakename == "Peter Lake")
Paul.nutrients.surface <- filter(PeterPaul.nutrients.surface, lakename == "Paul Lake")
```

# Run a Mann-Kendall test
```{r}
library(trend)
Peter.nutrients.surface$tp_ug<-as.numeric(Peter.nutrients.surface$tp_ug)
mk.test(Peter.nutrients.surface$tp_ug)
```
z=4.39, so We see a positive trend over time
z is the direction and magnitude of the trend
p=1.099e-0.5 so we reject the null hypothesis


However, it looks like there might be a breakpoint in our dataset. Further, we know that Peter Lake underwent experimental fertilization starting in May 1993, a perturbation which we might expect to have induced a regime shift in the ecosystem. In this case, we might want to find out whether there is a breakpoint, or changepoint, in our dataset. 


### Pettitt's Test
Pettitt's test is also included in the `trend` package. This nonparametric test will determine whether there is a shift in the central tendency of the time series and will tell us at what point the changepoint occurs (if it detects one). Note: Pettitt's Test will only test for one changepoint, and further tests must be run if multiple change points are suspected.

```{r}
# Test for change point
pettitt.test(Peter.nutrients.surface$tp_ug)

```
Is our p-value <0.05? IF SO THEN OUR CHANGE POINT IS SIGNIFICANT
It is 4.92e-09, so we;ll have to go into our data set and look at where change point happened
change point is 35, so scroll down to place 35 in data set. It happened in May 1993. 

# Run separate Mann-Kendall for each change point
```{r}
mk.test(Peter.nutrients.surface$tp_ug[1:34])
mk.test(Peter.nutrients.surface$tp_ug[35:132])

```


# Is there a second change point?
```{r}
pettitt.test(Peter.nutrients.surface$tp_ug[35:132])  ###1993 onwards
```
###35+79=114, so look at 114 row in datatable to see second change point. 


# Run another Mann-Kendall for the second change point
####Now split dataset into three pieces
```{r}
mk.test(Peter.nutrients.surface$tp_ug[35:113])   #where there was experimental fertilization
mk.test(Peter.nutrients.surface$tp_ug[114:132])  #not a significant trend 

```
###If z-score is positive, it's a positive trend. If z-score is negative, it is a negative trend
###There is not a significant trend for rows:114-132 becausw the p-value is 0.52, which is below 0.05. 
###Thus, there was increasing trend in phosphorus over time from 35:113

# Run the same test for Paul Lake. 
```{r}
mk.test(Paul.nutrients.surface$tp_ug)
pettitt.test(Paul.nutrients.surface$tp_ug)
```
###We have a non significant trend-p=0.15, z--1.43, so there is a negative trend in phosphorus


Petitts test:
p-value=0.12, change point=58
There is no change over time, and no detectable break point-so there's a flat trend over time in our reference lake.

# Add vertical lines to the original graph to represent change points in Peter Lake
```{r}
ggplot(PeterPaul.nutrients.surface, aes(x = sampledate, y = tp_ug, color = lakename)) + 
  geom_point() +
  scale_color_manual(values = c("#7fcdbb", "#253494"))+
  geom_vline(xintercept=as.Date("1993-05-20"),color="253494", origin= "1970-01-01", lty=2) +
  geom_vline(xintercept=as.Date("1998-01-01"), color="253494", lty=2)
###Choose a spot between the change points between 34 and 35
```
###How to specify origin as 1-1-1970 is above

###The Mann-Kendall test assumes all of your data points are independent of each other, which is not the case for this data


### Seasonal Mann-Kendall

Like a **Mann-Kendall Test**, the **Seasonal Mann-Kendall Test**, or **Hirsch-Slack Test**, analyzes trends in response variables over time. It replaces the traditional Mann-Kendall when there are seasonal trends in a dataset that obscure the overall direction of the trend. It is important to note that "seasonal" does not necessarily equate to actual seasons but can represent any time period within which there are oscillating temporal trends. The test needs at least two distinct seasons to operate.

For instance, we might want to know whether there is a change in discharge of the Eno River over the last 10 years.
```{r}
# Wrangle the USGS dataset
USGS.flow.data.trimmed <- USGS.flow.data %>%
  select(datetime, discharge.mean) %>%
  filter(datetime > as.Date("2008-12-31") & datetime < as.Date("2019-01-01")) ###Don't have to mess with dates b/c we're only using later values. Picking dates between 2009 through 2018
```

# Visualize the data
```{r}
ggplot(USGS.flow.data.trimmed, aes(x = datetime, y = discharge.mean)) +
  geom_point(size = 0.5, alpha = 0.5)
```
###What tempiraltrends in this data:
###Definite seasonality-peaks occurring during winter seasons
###There are outliers due to flood events
But in general, we see baseflow, which are the dark concentrated points by 0 axis
Seasonal mann-kendall tests can't have NA's 


### Interpolation
Some situations may require us to predict values for data points that fall within the time frame of our analyses but were not sampled. For instance, the `smk.test` function needs to take a time series format rather than a data frame, which cannot have any NAs. In this case, we will want to make an estimate of the missing values based on what we know about the dataset using a method called **interpolation.** There are several options for interpolation: 

  + **Means interpolation:** Defines values between sampled values as the mean value within a dataset. Uses the R function `aggregate`.
  =Takes mean value of data set and assign missing data points as the mean of the data set
  
  + **Piecewise constant interpolation:** Defines values between sampled values as the value of the nearest sampled value. Uses the R function `approx` with `method = "constant"`
=Results in jumps occurring between each  data point
  
  + **Linear interpolation:** Defines values between sampled values based on the slope between sampled values. Uses the R function `approx` with `method = "linear"`

  + **Spline interpolation:** Defines values between sampled values based on polynomial functions between sampled values and chooses the polynomials so that they fit smoothly together. Uses the R function `splinefun`.
=Looks like a linear interpolation but it smooths the line


+scale_y_log10 sets your y-axis from 1, 10, 100, 1000, etc 

Question: Under what circumstances would you consider each of these options for interpolation? 

> ANSWER: 
If we're looking at outliers, we'd use the linear or spline interpolations because the discharge variable is continuous and we'd want to connect the points.
If we measure water on a tuesday, we'd anticipate it would be in between monday's value and or wednesday's value
Spine ingerpolation is a polynomial, so in general the linear interpolation works better for a lot of environmental data


Tip: Check your dataset to see if there is an NA value in the first row. You may need to add a value for that first row or trim the dataset so that the new first row corresponds to the first measurement.

```{r}
# Run a linear interpolation of the dataset to fill in gaps
USGS.flow.data.interpolated <- approx(USGS.flow.data.trimmed$datetime,
                                      USGS.flow.data.trimmed$discharge.mean, 
                                      method = "linear", n = 3630)

###approximates datetime with the discharge mean
##tell it method=linear
###n=3630 -specify number of data points 
###this creates a list, so we have to turn it back into a data frame
```

# Turn the interpolated dataset into a proper dataframe
```{r}
USGS.flow.data.interpolated <- do.call(cbind.data.frame, USGS.flow.data.interpolated) ##gets us back to data frame
names(USGS.flow.data.interpolated) <- c("Date", "Discharge")   ###have to rename the columns
USGS.flow.data.interpolated$Date <- as.Date(USGS.flow.data.interpolated$Date, ##have to respecify the Date column as a date. Need to specify the origin too. 
                                            origin = "1970/01/01")
```



Mann-Kandall only takes a time series object, so convert data frame into a time series object
# Create a time series object
```{r}
USGS.flow.data.timeseries <- ts(USGS.flow.data.interpolated$Discharge, 
                                start = c(2009, 1) ,frequency = 12)

###Telling it to go in a frequency of 12 months per year (monthly), and start in the first month of 2009
###This generates a time series object, which you can't click on
```


# Run a Seasonal Mann-Kendall test
```{r}
USGS.smktest <- smk.test(USGS.flow.data.timeseries)
USGS.smktest  ###gives us overall trend with a zscore of 7.6, p-value of 2.04e-14. So we do see an increasing trend over time becuse the zscore is high, but  we have a lot of varability(varsS)

summary(USGS.smktest)
```


Any season with one or more stars is a significant trend
Notice in the winter, we see the greatest increases in our flow rates, while in the spring, we're not seeing increases or decreases in discharge-no significant trend over time

Interpreting results of the Seasonal Mann-Kendall Test: 

  + Overall z score and p-value: test the alternative hypothesis that the true change in response variable over time is not equal to zero
  
  + Monthly z score and p-value: test the alternative hypothesis that the true change in response variable over time for a given month is not equal to zero
  
  + S: reports trend. A positive value indicates response variable increased over time, and a negative value indicates response variable decreased over time
  
Question: How would you interpret the results of the Seasonal Mann-Kendall test for this example?

> ANSWER: 