---
title: "9: Data Visualization"
author: "Environmental Data Analytics | Kateri Salk"
date: "Spring 2019"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---

## LESSON OBJECTIVES
1. Perform simple data visualizations in the R package `ggplot`
2. Develop skills to adjust aesthetics and layers in graphs
3. Apply a decision tree framework for appropriate graphing methods

## SET UP YOUR DATA ANALYSIS SESSION

```{r, message = FALSE, warning = FALSE}
setwd("~/Desktop/Environmental Data Analytics/Environmental_Data_Analytics/Data/Processed")
library(tidyverse)

PeterPaul.chem.nutrients <- read.csv("NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv")
PeterPaul.nutrients.gathered <- read.csv("NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed.csv")
PeterPaul.chem.nutrients.summaries <- read.csv("NTL-LTER_Lake_Summaries_PeterPaul_Processed.csv")
EPAair <- read.csv("EPAair_O3_PM25_NC1718_Processed.csv")
attach(EPAair)
attach(PeterPaul.chem.nutrients)

```


##Unite Y, m, and d columns into a "date" column
###I fixed Assignment 4 and used the mutate function instead of separate so it kept the original complete date column in the EPAair datatable
```{r}
###EPAair <- EPAair %>%
 ### unite(date, Y, m, d)
###EPAair$date <- as.Date(EPAair$date, format="%m/%d/%y")


```

###unite(data, col, ..., sep = "_", remove = TRUE)


##Create a "month" column in EPAair datatable from the date column
```{r}
###EPAair<- mutate(EPAair, month = month(date))
```


##Create second EPAair data table by separating "date" column into Year, month, and day columns
```{r}
###EPAair2<- separate(EPAair, date, c("Y", "m", "d"))
```


##Create a new Date column in the PeterPaul.chem.nutrients data table in the proper date format from the SampleDate column
```{r}
setwd("~/Desktop/Environmental Data Analytics/Environmental_Data_Analytics/Data/Processed")
PeterPaul.chem.nutrients$Date<-as.Date(PeterPaul.chem.nutrients$sampledate, format="%m/%d/%y")

PeterPaul.chem.nutrients<- mutate(PeterPaul.chem.nutrients, month = month(Date))
PeterPaul.chem.nutrients<-mutate(PeterPaul.chem.nutrients, day=day(Date))


###PeterPaul.chem.nutrients<-separate(PeterPaul.chem.nutrients, Date, c("Y", "m", "d"))
```
###mutate() adds new variables and preserves existing variables.
###separate() turns a single character column into multiple columns but removes original column.



# GGPLOT

ggplot, called from the package `ggplot2`, is a graphing and image generation tool in R. This package is part of tidyverse. While base R has graphing capabilities, ggplot has the capacity for a wider range and more sophisticated options for graphing. 
ggplot has only a few rules: 

* The first line of ggplot code always starts with `ggplot()`
* A data frame must be specified within the `ggplot()` function. Additional datasets can be specified in subsequent layers.
* Aesthetics must be specified, most commonly x and y variables but including others. Aesthetics can be specified in the `ggplot()` function or in subsequent layers.
* Additional layers must be specified to fill the plot. 


ggplot(df, aes()) + geom

## Geoms
Create the actual graph
Here are some commonly used layers for plotting in ggplot: 

* geom_bar
* geom_histogram
* geom_freqpoly
* geom_boxplot
* geom_violin
* geom_dotplot
* geom_point
* geom_errorbar
* geom_smooth
* geom_line
* geom_area
* geom_abline (plus geom_hline and geom_vline)
* geom_text

## Aesthetics
Here are some commonly used aesthetic types that can be manipulated in ggplot: 

* color
* fill
* shape
* size
* transparency

time is always on x-axis

# Plotting continuous variables over time: Scatterplot

## Putting aesthetics in ggplot() function-DOESN'T WORK
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
ggplot(EPAair, aes(x = Date, y = Ozone)) + 
  geom_point()
```

##Putting aesthetics inside geom_point() function
```{r, fig.height = 3, fig.width = 4, warning = FALSE}
O3plot <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone))
print(O3plot)
```


# Fix this code

###MUST PUT aesthetics within geom_point() function 
###Can't put color= within aesthetics, so keep it within geom_point() but not in aesthetics!
###Geom_point can take a lot of information besides aesthetics; color is one of them

###Put aesthetics in geom_point
```{r}
O3plot2 <- ggplot(EPAair) +
  geom_point(aes(x = Date, y = Ozone), color = "Blue")
print(O3plot2)
```
###If you change color to "Ozone", the plot becomes a color gradient based on Ozone and adds a legend.




# Add additional variables

##Shape: Changes shape based on what year the data were collected

###Need to say as.factor(year) because the class of year is an integer, and we want it as a factor
###We can put color within aesthetics because we are assigning color by site name
```{r}
PMplot <- 
  ggplot(EPAair, aes(x = Date, y = PM2.5, shape = as.factor(year), color = Site.Name)) +
  geom_point()
print(PMplot)
```



# Separate plot with facets

###Faceting by Site.Name
```{r}
PMplot.faceted <-
  ggplot(EPAair, aes(x = month, y= PM2.5, shape = as.factor(year))) +
  geom_point() +
  facet_wrap(vars(Site.Name), nrow = 3)

print(PMplot.faceted)
```
###Faceting is used when you have one column with two or more "levels"



###Faceting by Year as rows and Site.Name as columns
```{r}
PMplot.faceted2 <-
  ggplot(EPAair, aes(x = month, y = PM2.5)) +
  geom_point() +
  facet_grid(year~Site.Name)
print(PMplot.faceted2)

```

###Faceting by Site.Name as rows and Years as columns
```{r}
PMplot.faceted3 <-
  ggplot(EPAair, aes(x = month, y = PM2.5)) +
  geom_point() +
  facet_grid(Site.Name~year)
print(PMplot.faceted3)

```


# Filter dataset within plot building

###For ex, if we just wanted to plot the "Blackstone" data, we would put Blackstone within quotation marks within subset function, even though in aesthetics function, you don't need quotations
```{r}
O3plot.Blackstone <- 
  ggplot(subset(EPAair, Site.Name == "Blackstone"), aes(x = Date, y = Ozone)) + 
  geom_point() +
  geom_line()
print(O3plot.Blackstone)
```
###Geom_line is connecting the data points along the x-axis. It's only useful if you want to see what the intervals between the data points look like. 



# Exercise: build your own scatterplots of PeterPaul.chem.nutrients

## 1. 


### Plot surface temperatures (depth=0) by day of year. 
### Color your points by year, and facet by lake in two rows.

```{r}
require(ggplot2)
PeterPaulScatterplot<-ggplot(subset(PeterPaul.chem.nutrients, depth=0), aes(x=day, y=temperature_C)) + geom_point(color=year4) +facet_wrap(vars(lakename), nrow=2) 
print(PeterPaulScatterplot)
```


```{r}
PeterPaulScatterplot<-ggplot(subset(PeterPaul.chem.nutrients, depth=0)) + geom_point(aes(x=day, y=temperature_C), color=year4) +facet_wrap(vars(lakename), nrow=2) 
print(PeterPaulScatterplot)
```



#2. 
## Plot temperature by date. Color your points by depth.
## Change the size of your point to 0.5

```{r}
require(ggplot2)
PeterPaulScatterplot2<-ggplot(PeterPaul.chem.nutrients, aes(x=Date, y=temperature_C,color=depth)) + geom_point(size=0.5) +facet_wrap(vars(lakename), nrow=2) 
print(PeterPaulScatterplot2)
```



# Plotting the relationship between two continuous variables: Scatterplot


```{r, fig.height = 3, fig.width = 4, warning = FALSE}
# Scatterplot
lightvsDO <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point()
print(lightvsDO)

# Adjust axes
lightvsDOfixed <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = irradianceWater, y = dissolvedOxygen)) +
  geom_point() +
  xlim(0, 250) +
  ylim(0, 20)
print(lightvsDOfixed)
###More light at the surface of the water; from 0-50 irradiance, there are low oxygen concentrations. 




# Depth in the fields of limnology and oceanography is on a reverse scale
###We normally consider dpeth as an independnet variable because it stays the same (temperature might vary)
###In limnology, depth is actually put on the y-axis
tempvsdepth <- 
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth)) +
  
  ggplot(PeterPaul.chem.nutrients, aes(x = temperature_C, y = depth, color = daynum)) +
  geom_point() +
  scale_y_reverse()
print(tempvsdepth)

###Scale_y_reverse() function 


NvsP <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = tn_ug, color = depth)) +
  geom_point() #+
  #geom_smooth(method = lm) +
  #geom_abline(aes(slope = 16, intercept = 0))
print(NvsP)

# Exercise: Plot relationships between air quality measurements

# 1. 
# Plot AQI values for ozone by PM2.5, colored by site. 
# Add a line of best fit for the linear regression of these variables.

  
```

### Plotting continuous vs. categorical variables

```{r, fig.height = 3, fig.width = 4, warning = FALSE}
 # Barplot + error bars
PeterPaul.nutrient.summaries <- PeterPaul.nutrients.gathered %>%
  group_by(lakename, nutrient) %>%
  summarise(sd = sd(concentration), 
            mean = mean(concentration))

Nutrientplot <- 
  ggplot(PeterPaul.nutrients.gathered) +
  geom_bar(aes(x = lakename, y = concentration, fill = as.factor(nutrient)), # why did we use fill?
           position = "dodge", stat = "summary", fun.y = "mean")             # what's happening here?
print(Nutrientplot)

Nutrientplot2 <- 
  ggplot(PeterPaul.nutrient.summaries, aes(x = lakename, y = mean, fill = as.factor(nutrient))) + #
  geom_bar(stat = "identity", position = "dodge") + # what does the stat command do?
  geom_errorbar(aes(ymin = mean-sd, ymax = mean+sd), # how do we specify error bars?
                 position = "dodge")
print(Nutrientplot2)

# Are there more effective ways to produce summary stats for categories?

# Box and whiskers plot
Nutrientplot3 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_boxplot(aes(color = nutrient)) # Why didn't we use "fill"?
print(Nutrientplot3)

# Dot plot
Nutrientplot4 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_dotplot(aes(color = nutrient), binaxis = "y", binwidth = 1, 
               stackdir = "center", position = "dodge") #
print(Nutrientplot4)

# Violin plot
Nutrientplot5 <-
  ggplot(PeterPaul.nutrients.gathered, aes(x = lakename, y = concentration)) +
  geom_violin(aes(color = nutrient)) #
print(Nutrientplot5)

# Frequency polygons
# Using a tidy dataset
Nutrientplot6 <-
  ggplot(PeterPaul.chem.nutrients) +
  geom_freqpoly(aes(x = tn_ug), color = "black") +
  geom_freqpoly(aes(x = tp_ug), color = "darkblue") +
  geom_freqpoly(aes(x = nh34), color = "darkgray") +
  geom_freqpoly(aes(x = no23), color = "gray") +
  geom_freqpoly(aes(x = po4), color = "blue") 
print(Nutrientplot6)

# Using a gathered dataset
Nutrientplot7 <-   
  ggplot(PeterPaul.nutrients.gathered) +
  geom_freqpoly(aes(x = concentration, color = nutrient))
print(Nutrientplot7)

# Exercise: Plot distributions of AQI values for EPAair

# 1. 
# Create a bar chart plus standard deviation error bars for PM2.5, divided by year. 
# Create separate bars for each site. 

# 2. 
# Create a new plot that better depicts the distribution of PM2.5 concentrations. 
# Divide your graph by year and site.

```
